---
title: "FinalProject_Team5_EDA"
output:
  pdf_document: default
  html_document: default
date: "2023-07-24"
---

```{r}

rm(list = ls()) # clear all variables in the workspace
library(ggplot2)
library(dplyr)
#library(lattice)
library(cowplot)
#library(reshape2)
library(RColorBrewer)
library(GGally)
library(car)
library(caret)
library(MASS)
library(moments)
library(stats)
library(corrplot)
library(lmtest)


# Convert the X and Y variables to categorical variables
forestfires = read.csv("forestfires.csv") 
forestfires$X = factor(forestfires$X)
forestfires$Y = factor(forestfires$Y)

# Divide the data set into training data and testing data, and the ratio between two subgroup data is 0.8:0.2.

#trainIndex = createDataPartition(forestfires$area, p = 0.8, list = FALSE)
#train_data = forestfires[trainIndex, ]
#test_data = forestfires[-trainIndex, ]
```

# Univariate analysis
## (A) Categorical data
```{r}

totalfires = nrow(forestfires) # total number of observations

boxcox(forestfires$area + 1 ~ 1, lambda = seq(-1, 0, 0.005))

# Generate the kernel density curve of burned area

par(mfrow = c(2,2))

xdata = forestfires$area
plot(density(xdata),xlab = "Area", ylab = "Density",main = "Kernal density curve of area", lwd = 2)
boxplot(xdata,xlab = "Area", ylab = "Density",main = "Boxplot of area", lwd = 2)
sk1 = skewness(xdata)
sprintf("Skewness of the area: %.2f",  sk1)

xdata2 = log(forestfires$area+1)
plot(density(xdata2),xlab = "Area", ylab = "Density",main = "Kernal density curve of log(area+1)", lwd = 2)
boxplot(xdata2,xlab = "Area", ylab = "Density",main = "Boxplot of log(area+1)", lwd = 2)
sk2 = skewness(xdata2)
sprintf("Skewness of the area: %.2f",  sk2)

boxcox(xdata+1~ 1, lambda = seq(-1, 0, 0.005))

```
## (A) Area vs. categorical variables

```{r}
# Create a new categorical variable representing the severity of the fires, based on the values of burned area
# No damage: area == 0
# Low damage: 0 < area < 10
# Moderate damage: 10 < area < 100
# High damage: area > 100

cut_points = c(-Inf, 0.00001, 10, 100, Inf)
severity_labels = c("No damage", "Low", "Moderate", "High")
severity = cut(forestfires$area, breaks = cut_points, labels = severity_labels, right = FALSE)
forestfires$severity = severity

# Group the data by X or Y coordinates
fires_X = forestfires %>%
  group_by(X,severity) %>%
  summarize(fires_count = n())
fires_X$severity = factor(fires_X$severity, levels = rev(severity_labels))

fires_Y = forestfires %>%
  group_by(Y,severity) %>%
  summarize(fires_count = n()) 
fires_Y$severity = factor(fires_Y$severity, levels = rev(severity_labels))

# Generate barplots of fires vs. X or fires vs. Y
par(mfrow=c(1,2))
h1 = ggplot(fires_X, aes(x = X, y = fires_count, fill = severity)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "X", y = "Count", title = "Fires vs. X") +
  scale_fill_discrete(name = "Severity")+
  scale_x_discrete(limits = c("1", "2", "3", "4", "5", "6", "7","8","9"),
                   labels = c("1", "2", "3", "4", "5", "6", "7","8","9"),) + 
  scale_y_continuous(name = "Count") +
  scale_fill_manual(values = brewer.pal(n = 7, name = 'RdGy'), name = "Severity") # use customized colors
h2 = ggplot(fires_Y, aes(x = Y, y = fires_count, fill = severity)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Y", y = "Count", title = "Fires vs. Y") +
  scale_fill_discrete(name = "Severity")+
  scale_x_discrete(limits = c("1", "2", "3", "4", "5", "6", "7","8","9"),
                   labels = c("1", "2", "3", "4", "5", "6", "7","8","9"),) +
  scale_y_continuous(name = "Count") +
  scale_fill_manual(values = brewer.pal(n = 7, name = 'RdGy'), name = "Severity") # use customized colors

# Group the data by month or day
severity_month <- forestfires %>%
  group_by(month,severity) %>%
  summarise(count = n())
severity_month$severity <- factor(severity_month$severity, levels = rev(severity_labels))
severity_day <- forestfires %>%
  group_by(day,severity) %>%
  summarise(count = n())
severity_day$severity <- factor(severity_day$severity, levels = rev(severity_labels))

# Generate barplots of fires vs. month
h3 = ggplot(severity_month, aes(x = month, y = count, fill = severity)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Month", y = "Count", title = "Fires in each month") +
  scale_fill_discrete(name = "Severity")+
  scale_x_discrete(limits = c("jan", "feb", "mar", 
                             "apr", "may", "jun", 
                             "jul","aug","sep",
                             "oct","nov","dec"),
                   labels = c("Jan", "Feb", "Mar", 
                             "Apr", "May", "June", 
                             "July","Aug","Sept",
                             "Oct","Nov","Dec"),) + # plot the data in chronological order
  scale_y_continuous(name = "Count",
                       sec.axis = sec_axis(~./totalfires, 
                                           name="Percentage")) +
  scale_fill_manual(values = brewer.pal(n = 7, name = 'RdGy'), name = "Severity") # use customized colors
# Generate barplots of fires vs. day
h4 = ggplot(severity_day, aes(x = day, y = count, fill = severity)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Day", y = "Count", title = "Fires in different day") +
  scale_fill_discrete(name = "Severity")+
  scale_x_discrete(limits = c("mon", "tue", "wed", 
                             "thu", "fri", "sat", 
                             "sun"),
                   labels = c("Mon", "Tue", "Wed", 
                             "Thu", "Fri", "Sat", 
                             "Sun"),) + # plot the data in chronological order
  scale_y_continuous(name = "Count",
                       sec.axis = sec_axis(~./totalfires, 
                                           name="Percentage")) +
  scale_fill_manual(values = brewer.pal(n = 7, name = 'RdGy'), name = "Severity") # use customized colors

plot_grid(h1, h2, h3, h4, labels = c( 'A', 'B','C','D'), label_size = 12) # Combine four plots to one figure




```
```{r}
# Create a new categorical variable called season with two levels: summer and others

forestfires$season = ifelse(forestfires$month %in% c("jun", "jul", "aug","sep"), "summer", "others")

severity_season <- forestfires %>%
  group_by(season,severity) %>%
  summarise(count = n())
severity_season$severity <- factor(severity_season$severity, levels = rev(severity_labels))

ggplot(severity_season, aes(x = season, y = count, fill = severity)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Season", y = "Count", title = "Fires in summer or other seasons") +
  scale_fill_discrete(name = "Severity")+
  scale_y_continuous(name = "Count",
                       sec.axis = sec_axis(~./totalfires, 
                                           name="Percentage")) +
  scale_fill_manual(values = brewer.pal(n = 7, name = 'RdGy'), name = "Severity") # use customized colors

```

## (B) Area vs. continous variable

### (i) Single variable

```{r}
realfires = forestfires[-which(forestfires$area==0),]
ghostfires = forestfires[which(forestfires$area==0),]

numerical_vars = forestfires[, sapply(forestfires, is.numeric)]
numerical_names = names(numerical_vars)
realfires_num = numerical_vars[-which(numerical_vars$area==0),]
ghostfires_num = numerical_vars[which(numerical_vars$area==0),]

par(mfrow = c(3, 3))
for (i in 1:(ncol(realfires_num)-1) )
{
  title1 = paste("KDE plot of " , numerical_names[i] )
  kde1 = density(realfires_num[,i])
  kde2 = density(ghostfires_num[,i])
  plot(kde1, main = title1, xlab = numerical_names[i], col = "black",lwd = 2)
  lines(kde2,col = "red", lwd = 2)
}


par(mfrow = c(3, 3))
for (i in 1:(ncol(realfires_num)-1) )
{
  title1 = paste("Box-Cox transformation of  " , numerical_names[i] )
  boxcox(realfires_num[,i]+1~ 1, lambda = seq(-1,2))
}

par(mfrow = c(3, 3))
for (i in 1:(ncol(realfires_num)-1))
{
  title2 = paste("Box plot of " , numerical_names[i] )
  boxdata1 = realfires_num[,i]
  boxplot(boxdata1, main = title2, xlab = numerical_names[i], lwd = 2)

}

par(mfrow = c(3, 3))
for (i in 1:(ncol(ghostfires_num)-1))
{
  title2 = paste("Box plot of " , numerical_names[i] )
  boxdata2 = ghostfires_num[,i]
  boxplot(boxdata2, main = title2, xlab = numerical_names[i], lwd = 2)
}

# Remove the data point with the z-score of FFMC greater than 3
z_scores_ffmc = scale(realfires$FFMC)
realfires = realfires[abs(z_scores_ffmc) <= 3,]
plot(density(realfires$FFMC))

# Four data points have been removed and the kernel density curved has been plotted in subplot-9
```
## (ii) area vs. numerical variables

```{r}
# Remove the outliers based on the values of FFMC
z_scores_ffmc = scale(forestfires$FFMC)
forestfires = forestfires[abs(z_scores_ffmc) <= 3,]
plot(density(realfires$FFMC))

forestfires_names = names(forestfires)
data_copy = forestfires
data_copy$severity <- factor(data_copy$severity, levels = rev(severity_labels)) # sort out the data according to the severity of fires

# Plot log-transformed area vs. individual continous variables
for (x_var in numerical_names) {
  if ( x_var == 'area' || x_var == "rain"){
    next
  }
  else{
ggplot(data_copy, aes(x =  get(x_var), y = log(area+1), color = severity)) +
    geom_point() +
    labs(title = paste("Log-transformed area vs ", x_var),x = x_var,y = "Log-transformed area") +
    scale_color_manual(values = brewer.pal(n = 7, name = 'RdGy'), name = "Severity") # use customized colors
}
}

# Generate the correlation map
CorrMatrix = cor(numerical_vars)
corrplot(CorrMatrix, method = 'color')
for (i in 1:nrow(CorrMatrix)) {
  for (j in 1:ncol(CorrMatrix)) {
    cov_val = round(CorrMatrix[i, j],2)
    text(i, 10-j, labels = cov_val, col = "black", cex = 0.8)
  }
}

```

# Section 3 Methodology

## 3.1 Simple model: multiple linear regression

```{r}

# Divide the realfires data into train and test data
trainIndex = createDataPartition(realfires$area, p = 0.8, list = FALSE)
train_data = realfires[trainIndex, ]
test_data = realfires[-trainIndex, ]

```
```{r}
# Fit a linear regression to the log_transformed area with all unchanged predictors 
# dataset = train_data

model_full_all = lm(log(area+1) ~.-severity-season , data = train_data)
summary(model_full_all)
anova(model_full_all)
par(mfrow = c(2,2))
plot(model_full_all)

boxcox(model_full_all)

model_full_all02 = lm(log(area+1)^0.5 ~.-severity-season , data = train_data)
summary(model_full_all02)
anova(model_full_all02)
par(mfrow = c(2,2))
plot(model_full_all02)

bptest(model_full_all)
```

Although box-cox suggests a square-root transformation of the response, we did found adequate evidence of non-normality variance,. Therefore, we will discard the square-root transformation and keep the log-transformed area at this moment.

```{r}
#rm(model_full_all02)
# check the collinearity of predictors
vif_output = vif(model_full_all)
print(vif_output)

```

Looking at the output of vif function, we can read that the variable DC has a high generalized vif, indicating a high correlation of it with other predictors in the model. Let's eliminate this variable to refit a model and compare the reduced model with the full model.

```{r}
model_reduced = lm(log(area+1)~.- severity-season-DC, data = train_data)
anova(model_reduced,model_full_all)
```

Here, it looks like we have a contradictory conclusion and we can't fix the collinearity problem by simply discarding the variable. We will have to consider otehr methods like principle component analysis (PCA), ridge regression and etc.

Now, let's fix the skewness of FFMC and RH
```{r}
par(mfrow = c(1,2))
boxcox(train_data$RH~1)
boxcox(train_data$FFMC~1,lambda = c(-20,20))

skewness(train_data$FFMC)
skewness(sqrt(train_data$FFMC))
skewness(log(train_data$FFMC))
skewness(train_data$FFMC^10)

plot(density(train_data$FFMC^10))

model_full_all02 = lm(log(area+1)~.-severity-season+log(RH) ,data = train_data)
anova(model_full_all,model_full_all02)

model_full_all03 = lm(log(area+1)~.-severity-season-RH+log(RH) ,data = train_data)
model_full_all04 = lm(log(area+1)~.-severity-season-RH+ I(1/RH) ,data = train_data)
model_full_all05 = lm(log(area+1)~.-severity-season+ I(1/RH) ,data = train_data)
anova(model_full_all05,model_full_all)

summary(model_full_all03)$r.square
summary(model_full_all04)$r.square
summary(model_full_all)$r.square

model_full_all06 = lm(log(area+1) ~.- severity-season-RH+I(1/RH) - FFMC+ poly(FFMC, degree = 20), data = train_data)

model_updated = model_full_all04 # the model that has best performance is the model that replace RH with 1/RH
```

```{r}
rm(model_full_all02)
rm(model_full_all03)
rm(model_full_all04)
rm(model_full_all05)
rm(model_full_all06)

prediction = predict(model_updated,newdata = test_data)

plot(log(test_data$area+1),prediction,xlim = c(0,6),ylim = c(0,6))
lines(0:6, 0:6, type = "l", col = "red")

rmse = sqrt(mean((log(test_data$area+1) - prediction)^2))
rmse

```
# transform the numerical variable according to the results of box-cox 

for (i in 1:(ncol(numerical_vars)-1) )
{
  title1 = paste("Box-Cox transformation of  " , numerical_names[i] )
  boxcox(numerical_vars[,i]+1~ 1, lambda = seq(0,20))
}
FFMC = forestfires$FFMC
DMC = forestfires$DMC^0.5
DC = forestfires$DC^1.5
ISI = forestfires$ISI^0.5
temp = forestfires$temp
RH = log(forestfires$RH)
wind = forestfires$wind^0.5
area = forestfires$area
X = forestfires$X
Y = forestfires$Y
season = forestfires$season

fires_boxcox = data.frame(FFMC = FFMC, DMC = DMC, DC = DC, ISI = ISI, temp = temp,
                          RH = RH, wind = wind, area = area, X = X, Y = Y, season = season)
nonzero_boxcox = fires_boxcox[-which(fires_boxcox$area==0),]


# The full model that include all non-transformed variables
model1 = lm(log(area+1)^0.5 ~. -severity-season, data = realfires)
summary(model1)
par(mfrow =c(2,2))
plot(model1)
anova(model1)

```
```{r}
model2 = lm(log(area+1)^0.5 ~. -severity-season  -day, data = nonzero)
summary(model2)
anova(model2,model1)


model3 = lm(log(area+1)^0.5 ~. -severity-season-day  -month+season, data = nonzero)
summary(model3)
anova(model3,model1)
```



# Multivariate analysis

```{r}
#ggplot( ) +
 # geom_point(alpha = 0.7) +  # Use alpha to adjust point transparency (optional)
 # geom_smooth(method = 'lm', se = FALSE, color = 'black', alpha = 0.2) +  # Add regression lines (optional)
  #ggpairs(forestfires, mapping = aes(color = severity,size = 0.01), 
  #      lower = list(continuous = wrap("points", alpha = 0.5)),
  #      upper = list(continuous = wrap("points", alpha = 0.5)),
  #      diag = list(continuous = wrap("densityDiag", alpha = 0.5)))+
  #scale_color_manual(values = brewer.pal(n = 9, name = 'RdGy'), name = "Severity") # use customized colors
plot(forestfires)
```



Preparing the data for modeling

```{r}

# Perform transformation on FFMC, ISI, rain and area to minimize the effect of ourliers

FFMC_log = log(forestfires$FFMC)
ISI_log = log(forestfires$ISI)
rain_log = log(forestfires$rain + 1) # +1 to avoid infinite log(rain)
area_log = log(forestfires$area + 1) # +1 to avoid infinite log(area)

log_frame = data.frame(FFMC_log= FFMC_log,ISI_log = ISI_log, rain_log= rain_log, area_log=area_log )
log_moment= descr(log_frame)
print(log_moment)

```

FFMC and rain are still having high skew and kurtosis values. We consider to remove the outliners in the two columns, using z-score method, to minimize the effect of outliers in the model.

```{r}




summary_table_removed  = descr(log_frame_removed)
print(summary_table_removed)

logname = names(log_frame)

par(mfrow = c(2, 2))
for (i in 1:4)
{
  title2 = paste("KDE plot of " , logname[i] )
  kde = density(log_frame_removed[,i])
  plot(kde, main = title2, xlab = logname[i], lwd = 2)
}

```

# Section 3.1 

# Simple Linear regression
Let's do nothing on the data and fit a simple linear model. 

```{r}
reduced_data = forestfires[,-which(names(forestfires)=="severity")]
varnames = names(reduced_data);

fullmodel = lm( area +1 ~., data = reduced_data)

model_summary = summary(fullmodel)
fstat <- model_summary$fstatistic["value"]
df1 <- model_summary$fstatistic["numdf"]
df2 <- model_summary$fstatistic["dendf"]
pvalue = 1 - pf(fstat, df1, df2)

boxcox_model = lm( (2-2/sqrt(area+1)) ~., data = reduced_data)
#boxcox_model = lm( sqrt(area + 1) ~., data = reduced_data)
summary(boxcox_model)
plot(boxcox_model)

#robust_model = lmrob(area +1 ~., data = reduced_data)
#summary(robust_model)
#plot(robust_model)

```

```{r}
rm(modeltest)
pvalues = rep(0, times = length(varnames))
modeltest = data.frame(name = varnames, pvalues = pvalues)
i = 1

for (x_var in varnames) {
  if (x_var == "area"){
    next
  }
  
  data_copy = reduced_data[, -which(names(reduced_data) == x_var)]
  
  reduced_model = lm( log(area+1)~., data = data_copy)
  results = anova(reduced_model,fullmodel)
  modeltest$pvalues[i] = results$`Pr(>F)`[2]
  i = i+1
}

# remove the variable from the data set that has the highest p-value
idx_maxp = which.max(modeltest$pvalues)
reduced_data = reduced_data[,-idx_maxp]
varnames = varnames[-idx_maxp]

fullmodel  = lm(log(area+1) ~., data = reduced_data)
model_summary = summary(fullmodel)
f_statistic <- model_summary$fstatistic["value"]
df1 <- model_summary$fstatistic["numdf"]
df2 <- model_summary$fstatistic["dendf"]
p_value_f_test = 1 - pf(f_statistic, df1, df2)


```



```{r}
plot(slr_allvar,which = 1)
bptest(slr_allvar)
#linearity_mod = lm(abs(slr_allvar$residuals)~slr_allvar$fitted.values  + I(slr_allvar$fitted.values^2) + I(slr_allvar$fitted.values^4))
#summary(linearity_mod)
```

According to the plot, with the presence of outliars, we concluded the homoscedasticify of the data.


```{r}
# normality check 
plot(slr_allvar,which = 2)

# error correlation check
dwtest(slr_allvar)

# model collinearity
round(vif(slr_allvar), dig = 2)

```

```{r}
var_names = names(forestfires)

for (i in 1:12){
  if (is.numeric(forestfires[,i]))
  {
    title3 = paste("Residuals vs. " , var_names[i])
    plot(forestfires[,i], slr_allvar$residuals, main = title3, xlab = var_names[i], ylab = "Residuals", lwd = 2)
  }
  else {
    next
  }
    
}
```

```{r}

fullmodel = lm(area~., data= train_data)

nummodel = lm(area~.-month-day,data = train_data)

nummodel2 = lm(area~.-X-Y,data = train_data)

anova(nummodel,fullmodel)
anova(nummodel2,fullmodel)

#areapca= prcomp(train_data[,-13])
#round(areapca$sdev,3)[1:50]

```